---
title: "Embat FinScore — Exploratory Data Analysis"
author: "Generated with Claude Code"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 5.5
)

pkgs <- c("data.table", "dplyr", "ggplot2", "scales", "lubridate", "skimr", "naniar", "stringr", "DBI", "duckdb")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")

library(data.table)
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)
library(skimr)
library(naniar)
library(stringr)
library(DBI)
library(duckdb)

theme_set(theme_minimal(base_size = 11))

data_dir <- "../data"

# Base DuckDB generada con: python analysis/build_db.py
db_path <- file.path(data_dir, "embat.duckdb")
if (!file.exists(db_path)) stop("Falta ", db_path, ". Ejecuta `python analysis/build_db.py` desde la raíz del proyecto.")
con <- dbConnect(duckdb::duckdb(), dbdir = db_path, read_only = TRUE)
q <- function(sql) as.data.table(dbGetQuery(con, sql))
```

# Overview

Synthetic dataset: **1,286 companies** across **250 business groups**, 24 months of financial
history (2024-09-01 to 2026-09-01). See `data/data_dictionary.md` for the full schema.

Files:

| file | rows (approx) | notes |
|---|---|---|
| groups.csv | 250 | 1 row / group |
| companies.csv | 1,286 | 1 row / company |
| banking_products.csv | 5,987 | bank accounts, cards, etc. |
| debt_products.csv | 2,239 | loans, leasing, factoring, etc. |
| debt_schedule_config.csv | 87 | amortization terms |
| balances.csv | 7,996 | balance snapshot as of 2026-09-01 |
| transactions.csv | 2,556,437 | **large (~450MB)** — bank transactions |
| invoices.csv | 897,894 | **large (~165MB)** — ERP invoices |

All tables are read from the local DuckDB database (`data/embat.duckdb`, built from the CSVs with
`python analysis/build_db.py`) and converted to `data.table`.

---

# 1. Groups

```{r groups-load}
groups <- q("SELECT * FROM groups")
str(groups)
skim(groups)
```

```{r groups-eda}
groups[, .N, by = erp][order(-N)]

ggplot(groups, aes(x = n_companies_in_sample)) +
  geom_histogram(binwidth = 1, fill = "#4C72B0") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(title = "Group size distribution (companies per group)",
       x = "n_companies_in_sample", y = "Count of groups")

summary(groups$n_companies_in_sample)
```

---

# 2. Companies

```{r companies-load}
companies <- q("SELECT * FROM companies")
companies[, created_at := as_datetime(created_at)]
str(companies)
```

```{r companies-missing}
gg_miss_var(as.data.frame(companies), show_pct = TRUE) +
  labs(title = "Missingness by column — companies.csv")
```

```{r companies-eda}
companies[, .N, by = country][order(-N)]
companies[, .N, by = currency][order(-N)]
companies[, .N, by = erp][order(-N)]

ggplot(companies, aes(x = as_date(created_at))) +
  geom_histogram(binwidth = 30, fill = "#55A868") +
  labs(title = "Company onboarding over time", x = "created_at", y = "New companies (30-day bins)")

# companies per group, cross-check against groups.csv
comp_per_group <- companies[, .N, by = group_id]
summary(comp_per_group$N)
```

---

# 3. Banking products (accounts, cards, etc.)

```{r banking-load}
banking <- q("SELECT * FROM banking_products")
banking[, created_at := as_datetime(created_at)]
str(banking)
```

```{r banking-eda}
banking[, .N, by = type][order(-N)]

ggplot(banking, aes(y = reorder(type, type, length))) +
  geom_bar(fill = "#4C72B0") +
  labs(title = "Banking product type distribution", y = NULL, x = "Count")

top_banks <- banking[, .N, by = bank_name][order(-N)][1:15]
ggplot(top_banks, aes(x = N, y = reorder(bank_name, N))) +
  geom_col(fill = "#C44E52") +
  labs(title = "Top 15 banks by product count", x = "Count", y = NULL)

banking[, .N, by = currency][order(-N)]

ggplot(banking, aes(x = as_date(created_at))) +
  geom_histogram(binwidth = 30, fill = "#8172B2") +
  labs(title = "Banking product creation over time", x = "created_at", y = "Count (30-day bins)")
```

---

# 4. Debt products (loans, leasing, factoring, etc.)

```{r debt-load}
debt <- q("SELECT * FROM debt_products")
debt[, created_at := as_datetime(created_at)]
str(debt)
```

```{r debt-eda}
debt[, .N, by = type][order(-N)]

ggplot(debt, aes(y = reorder(type, type, length))) +
  geom_bar(fill = "#4C72B0") +
  labs(title = "Debt product type distribution", y = NULL, x = "Count")

# granted / outstanding are often negative (liabilities); look at magnitude
debt_amt <- melt(debt, id.vars = c("product_id", "type"),
                  measure.vars = c("granted", "outstanding", "liquidity"),
                  variable.name = "metric", value.name = "value", na.rm = TRUE)

ggplot(debt_amt, aes(x = abs(value))) +
  geom_histogram(bins = 60, fill = "#DD8452") +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Debt amounts by metric (log10 abs scale)", x = "abs(value)", y = "Count")

debt_amt[, .(median = median(value), mean = mean(value), n = .N), by = .(type, metric)]

# outstanding vs granted utilization, by type
util <- debt[!is.na(granted) & !is.na(outstanding) & granted != 0]
util[, utilization := outstanding / granted]
ggplot(util, aes(x = type, y = utilization)) +
  geom_boxplot(outlier.alpha = 0.2, fill = "#4C72B0") +
  coord_cartesian(ylim = c(-1, 2)) +
  labs(title = "Outstanding / granted ratio by debt type", x = NULL, y = "utilization")
```

---

# 5. Debt schedule config (amortization terms)

```{r schedule-load}
schedule <- q("SELECT * FROM debt_schedule_config")
schedule[, `:=`(next_payment_date = as_datetime(next_payment_date),
                 last_payment_date = as_datetime(last_payment_date))]
str(schedule)
```

```{r schedule-eda}
schedule[, .N, by = amortization_type]
schedule[, .N, by = interest_calc_method]
schedule[, .N, by = amortising_frequency]
schedule[, .N, by = interest_type]

ggplot(schedule, aes(x = annual_interest_rate_or_spread, fill = interest_type)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.6) +
  labs(title = "Interest rate / spread distribution", x = "annual_interest_rate_or_spread")

ggplot(schedule, aes(x = total_periods)) +
  geom_histogram(bins = 30, fill = "#55A868") +
  labs(title = "Total periods (installments) distribution")

schedule[, pct_repaid := 1 - outstanding_balance / granted_balance]
summary(schedule$pct_repaid)
```

---

# 6. Balances (snapshot as of 2026-09-01)

```{r balances-load}
balances <- q("SELECT * FROM balances")
balances[, date := as_datetime(date)]
str(balances)
```

```{r balances-missing}
gg_miss_var(as.data.frame(balances), show_pct = TRUE) +
  labs(title = "Missingness by column — balances.csv")
```

```{r balances-eda}
summary(balances$balance)

ggplot(balances[abs(balance) > 1], aes(x = balance)) +
  geom_histogram(bins = 80, fill = "#4C72B0") +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(base = 10),
    labels = scales::comma
  ) +
  labs(title = "Balance distribution (pseudo-log x-scale, |balance| > 1)")

# join to product type for context
bal_typed <- merge(balances,
                    rbind(banking[, .(product_id, type, label_source = "banking")],
                          debt[, .(product_id, type, label_source = "debt")]),
                    by = "product_id", all.x = TRUE)

bal_typed[, .(median_balance = median(balance), n = .N), by = type][order(-n)]

ggplot(bal_typed[!is.na(type)], aes(x = type, y = balance)) +
  geom_boxplot(outlier.alpha = 0.15, fill = "#DD8452") +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(base = 10),
    labels = scales::comma
  ) +
  labs(title = "Balance by product type (pseudo-log y-scale)", x = NULL)
```

---

# 7. Transactions (large file — 2.5M rows)

```{r tx-load}
tx <- q("SELECT * FROM transactions")
tx[, `:=`(date = as_datetime(date), value_date = as_datetime(value_date))]
cat("Rows:", format(nrow(tx), big.mark = ","), " | Cols:", ncol(tx), "\n")
str(tx)
```

```{r tx-missing}
tx_missing_summary <- sapply(tx, function(col) mean(is.na(col) | (is.character(col) & col == "")))
data.table(column = names(tx_missing_summary), pct_missing = round(tx_missing_summary * 100, 2))[order(-pct_missing)]
```

```{r tx-volume-over-time}
tx[, month := floor_date(date, "month")]
tx_monthly <- tx[, .(n_tx = .N, total_amount = sum(amount)), by = month][order(month)]

ggplot(tx_monthly, aes(x = month, y = n_tx)) +
  geom_col(fill = "#4C72B0") +
  labs(title = "Transaction volume by month", x = NULL, y = "Number of transactions")

ggplot(tx_monthly, aes(x = month, y = total_amount)) +
  geom_col(fill = "#C44E52") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Net transaction amount by month", x = NULL, y = "Sum(amount)")
```

```{r tx-amount-dist}
summary(tx$amount)

ggplot(tx[abs(amount) > 1], aes(x = amount)) +
  geom_histogram(bins = 100, fill = "#8172B2") +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(base = 10),
    labels = scales::comma
  ) +
  labs(title = "Transaction amount distribution (pseudo-log x-scale)")

tx[, .(n = .N), by = .(inflow = amount > 0)][, pct := n / sum(n)][]
```

```{r tx-status-category}
tx[, .N, by = status][order(-N)]
tx[, .N, by = accounting_status][order(-N)]

top_cat <- tx[, .N, by = category][order(-N)][1:20]
ggplot(top_cat, aes(x = N, y = reorder(category, N))) +
  geom_col(fill = "#55A868") +
  labs(title = "Top 20 transaction categories", x = "Count", y = NULL)
```

```{r tx-counterparty}
tx[, has_counterparty := !is.na(counterparty_id)]
tx[, .N, by = has_counterparty][, pct := N / sum(N)][]

top_counterparties <- tx[has_counterparty == TRUE, .N, by = counterparty_id][order(-N)][1:15]
top_counterparties
```

```{r tx-per-company}
tx_per_company <- tx[, .(n_tx = .N, net_amount = sum(amount)), by = company_id]
summary(tx_per_company$n_tx)

ggplot(tx_per_company, aes(x = n_tx)) +
  geom_histogram(bins = 60, fill = "#4C72B0") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Transactions per company (log10 x-scale)")
```

---

# 8. Invoices (large file — ~900K rows)

```{r inv-load}
inv <- q("SELECT * FROM invoices")
inv[, `:=`(issuance_date = as_datetime(issuance_date),
           due_date = as_datetime(due_date),
           payment_date = as_datetime(payment_date))]
cat("Rows:", format(nrow(inv), big.mark = ","), " | Cols:", ncol(inv), "\n")
str(inv)
```

```{r inv-missing}
inv_missing_summary <- sapply(inv, function(col) mean(is.na(col) | (is.character(col) & col == "")))
data.table(column = names(inv_missing_summary), pct_missing = round(inv_missing_summary * 100, 2))[order(-pct_missing)]
```

```{r inv-doctype-status}
inv[, .N, by = document_type][order(-N)]
inv[, .N, by = status][order(-N)]
inv[, .N, by = currency][order(-N)][1:10]
```

```{r inv-volume-over-time}
inv[, month := floor_date(issuance_date, "month")]
inv_monthly <- inv[, .(n_inv = .N, total_amount = sum(amount, na.rm = TRUE)), by = month][order(month)]

ggplot(inv_monthly, aes(x = month, y = n_inv)) +
  geom_col(fill = "#4C72B0") +
  labs(title = "Invoice volume by month (issuance_date)", x = NULL, y = "Number of invoices")
```

```{r inv-amounts}
summary(inv$amount)
summary(inv$pending_amount)

ggplot(inv[abs(amount) > 1], aes(x = amount)) +
  geom_histogram(bins = 100, fill = "#DD8452") +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(base = 10),
    labels = scales::comma
  ) +
  labs(title = "Invoice amount distribution (pseudo-log x-scale)")

# payment behaviour: days late = payment_date - due_date, paid invoices only
inv_paid <- inv[status == "paid" & !is.na(payment_date) & !is.na(due_date)]
inv_paid[, days_late := as.numeric(difftime(payment_date, due_date, units = "days"))]

summary(inv_paid$days_late)

ggplot(inv_paid[abs(days_late) < 120], aes(x = days_late)) +
  geom_histogram(binwidth = 2, fill = "#4C72B0") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Days late (payment_date - due_date), paid invoices, clipped to +/-120d",
       x = "Days late", y = "Count")
```

```{r inv-per-company}
inv_per_company <- inv[, .(n_inv = .N, net_amount = sum(amount, na.rm = TRUE)), by = company_id]
summary(inv_per_company$n_inv)

ggplot(inv_per_company, aes(x = n_inv)) +
  geom_histogram(bins = 60, fill = "#55A868") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Invoices per company (log10 x-scale)")
```

---

# 9. Cross-file joins

```{r cross-joins}
# companies with number of banking/debt products, transactions, invoices, and net balance
company_summary <- companies[, .(company_id, group_id, country, currency)]

company_summary <- merge(company_summary, banking[, .(n_banking_products = .N), by = company_id],
                          by = "company_id", all.x = TRUE)
company_summary <- merge(company_summary, debt[, .(n_debt_products = .N), by = company_id],
                          by = "company_id", all.x = TRUE)
company_summary <- merge(company_summary, tx_per_company[, .(company_id, n_tx, net_tx_amount = net_amount)],
                          by = "company_id", all.x = TRUE)
company_summary <- merge(company_summary, inv_per_company[, .(company_id, n_inv, net_inv_amount = net_amount)],
                          by = "company_id", all.x = TRUE)
company_summary <- merge(company_summary, balances[, .(total_balance = sum(balance, na.rm = TRUE)), by = company_id],
                          by = "company_id", all.x = TRUE)

for (col in c("n_banking_products", "n_debt_products", "n_tx", "n_inv")) {
  company_summary[is.na(get(col)), (col) := 0]
}

head(company_summary, 15)
skim(company_summary)
```

```{r cross-correlations}
num_cols <- c("n_banking_products", "n_debt_products", "n_tx", "n_inv", "total_balance")
cor_mat <- cor(company_summary[, ..num_cols], use = "pairwise.complete.obs")
round(cor_mat, 2)

ggplot(company_summary, aes(x = n_banking_products, y = n_tx)) +
  geom_point(alpha = 0.3, color = "#4C72B0") +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Transaction count vs. number of banking products per company")
```

```{r cross-groups}
group_summary <- merge(groups, companies[, .(n_companies_actual = .N), by = group_id],
                        by = "group_id", all.x = TRUE)
group_summary[n_companies_in_sample != n_companies_actual]  # sanity check, should be empty/small
```

---

# 10. Summary of findings

Use this section to jot down the key takeaways once the notebook has been run end to end, e.g.:

- Data volume and time coverage per file
- Missingness hot-spots worth handling before modeling
- Skew/outliers in amount fields (log-scale needed almost everywhere)
- Relationship between activity (transactions/invoices) and portfolio size per company
- Any inconsistencies surfaced by the cross-file sanity checks in Section 9
